---
title: "Dietrich sinking velocity"
author: "David Kaiser"
date: "2018-02-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("Dietrich_velocity.R")
```

## Description

Calculates the sinking velocity of a particle in a fluid according to formulas by [Dietrich (1982)](http://onlinelibrary.wiley.com/doi/10.1029/WR018i006p01615/abstract). The function was originally written to calculate the theoretical sinking velocity of microplastic particles with diameters < 5 mm but > 200 µm. This calculation was used e.g. by [Kowalski et al. (2016)](https://www.sciencedirect.com/science/article/pii/S0025326X16303848), because the commonly used Stokes formula overestimates sinking velocity for particles with diameters > 200 µm. 
The formula by Dietrich considers effects of fluid denisty as well as particle size, shape and roundness. For non-spherical particles, the size/diameter can be expressed as equivalent spherical diameter (ESD) [(e.g. Kumar et al. 2010)](https://www.sciencedirect.com/science/article/pii/S0278434310003134).
the calculation only works when the density of the particle is higher than that of the fluid, i.e. when there is downward sinking. Otherwise the result is NaN.

The function requires the packages **marelac** and **matrixStats** to be installed. The packages are loaded by the function with a call to library().

## Arguments

* *salinity* -- in psu
* *temperature* -- in °C
* *depth* -- in m
* *latitude* -- in °N (negative for southern hemisphere), used to calculate pressure for density as well as gravity
* *particle.density* -- in kg m^-3^
* *particle.diameter* -- in m
* *powers.p* = 6 -- [Powers](https://pubs.geoscienceworld.org/sepm/jsedres/article-abstract/23/2/117/112811/a-new-roundness-scale-for-sedimentary-particles?redirectedFrom=fulltext) roundness; defaults to 6 for perfectly round areas, including shperes
* *CSF* = 1 -- [Corey Shape Factor](https://www.researchgate.net/publication/252625134_Settling_Velocities_of_Circular_Cylinders_at_Low_Reynolds_Numbers); defaults to 1 for spheres

## Value

A numeric value of the sinking velocity in m s^-1^

## Example

Calculate the sinking velocity [m s^-1^] for a polystyrene sphere (density is 1050 kg m^-3^) with a diameter of 1.5 mm diameter, in temperate ocean surface water.

```{r Example, message=FALSE, warning=FALSE, paged.print=FALSE}
Dietrich.velocity.m.sec(salinity = 36, 
                        temperature = 20, 
                        depth = 0.2, 
                        latitude = 40, 
                        particle.density = 1050, 
                        particle.diameter = 0.0015)
```

Sinking velocity cannot be calculated for particles with lower density than the fluid, e.g. polystyrene in temperate ocean surface water.

```{r Example2, message=FALSE, warning=FALSE, paged.print=FALSE}
Dietrich.velocity.m.sec(salinity = 36, 
                        temperature = 20, 
                        depth = 0.2, 
                        latitude = 40, 
                        particle.density = 955, 
                        particle.diameter = 0.0015)
```